Effectiveness of Instant Feedback on Developer Navigation Practices

### 2013-09-02 paper_id1.tex
monday nearly done with related work and pre-study

@@

Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.  We overcame this by dividing the study in to phases where the participants themselves form a control group.  The first phase baselines their activities determining the current state of their practices.  The second phase presents the intervention to the participants.  This phase the participants do not receive continuous feedback on their compliance with the intervention.  This phase simulates the traditional approach of communicating an intervention then returning with an audit to determine whether participants are complaint with it.   The last phases turns on continuous feedback to the participants on their compliance with the intervention.  As they receive the feedback, the participants can determine for themselves how much they are implementing the intervention.  In the last phase they possess complete knowledge of the intervention and their compliance with it making it plausible that they can continuously control their compliance.

@@

\subsection{Setup the practice}
Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.  In this study we selected the structured navigation practice shown by Robillard et. al. as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .  Studying structured navigation is a natural fit for capturing commands issued in the IDE.  The first step is to create the classification of commands that are involved in navigation.  Then differentiate which commands we consider as structured navigation from unstructured navigation.

@@

After classifying the commands we assigned points to the commands in the same configuration file.  Structured navigation commands received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.  When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.  Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{}.  Designing the points and levels such that above average developers can complete all levels during the study period is ideal.

@@

\section{Related Work}
%update and expand
Researchers have developed systems that aim to answer specific questions using data from command logs. Zorro \cite{V:Johnson2007Automated} leverages the HackyStat data to determine if developers are adhering to test-driven development.  The tool employs a similar approach dividing the work of the developer into episodes delineated by when test cases pass.  The researchers define multiple types of episodes then determine whether the developer uses Test-Driven-Development or not in each episode.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different approach to focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

>>>>

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering\cite{wbsnipes:Beecham2008Motivation}. It has some valuable lists of factors that motivate software engineers with the most common being "the work" motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.

\section{Assess the Viability of Gamification for Software Engineering}

\subsection{Assessment Background}
When discussing gamification, folks raise a concern about replacing an activity that is intrinsically movitated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.  The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.  Each question is ranked for its correlation to achievement motivation. Results indicate that some key questions about the diligence with which students approach their work are correlated with achievement motivation.  The questions in this questionnaire and their ranking may be useful for evaluating achievement motivation in further studies.  Take care when choosing appropriate questions that for the context of the survey.

ii.	Determine through a questionnaire survey how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers with recognition by badges and points. The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.

\subsection{Assessment Results}

Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?  We segmented resuults by country and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  The overall ABB response is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.

95\% of the developers surveyed are likely to try tools and practices recommended to them through an automated usage tracking system.  This indicates that a recommender system would positively impact the deployment of software engineering tools and practices at ABB.

One area we questioned was whether sharing this information would be a concern for developers.  A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.  Over 90\% of respondents are comfortable with sharing anonymous/non-anonymous information with their team members.  The figure below shows how this comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.  
Two conclusions from this feedback are that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents ranking in 1st or 2nd place.   The lowest was having badges posted on their You@ABB profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.  Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.

Developers were asked to rank awards for different actions by which they would find the most desirable.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.

Based on the survey results we draw the following conclusions for this project going forward:
•	95\% of developer responding would try tools and practices suggested by an automated recommender system particularly if they were provided in context with the current workflow pattern the developer follows.
•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
•	Over 90\% of respondents approved of sharing data with their team members

@@

Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.  We overcame this by dividing the study into phases where the participants themselves form a control group.  The first phase baselines their activities determining the current state of their practices.  The second phase presents the intervention to the participants.  This phase the participants do not receive continuous feedback on their compliance with the intervention.  This phase simulates the traditional approach of communicating an intervention then returning with an audit to determine whether participants are complaint with it.   The last phases turns on continuous feedback to the participants on their compliance with the intervention.  As they receive the feedback, the participants can determine for themselves how much they are implementing the intervention.  In the last phase they possess complete knowledge of the intervention and their compliance with it making it plausible that they can continuously control their compliance.

@@

\subsection{Setup the practice}

Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.  Companies are gameifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.  Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.

First it is important to consider the existing motivation people may have for performing the activity.  Applying gamefiication to activities that people are motivated intrinsically to do may decrease their motivation.  For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.    In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.   If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.

Several studies support the use of games as a learning aid or as the core part of a training program.  Naicker, Vincent and Adams \cite{wbsnipes:Amory1999Use} studied which type of computer game had the best response for enjoyment of play among undergraduate students.  The authors concluded that adventure games provide the best foundation for development of teaching resources.  Students rated sound, graphics, story line, and use of technology as important game elements.

Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.  In this study we selected the structured navigation practice shown by Robillard et. al. as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .  Studying structured navigation is a natural fit for capturing commands issued in the IDE.  The first step is to create the classification of commands that are involved in navigation.  Then differentiate which commands we consider as structured navigation from unstructured navigation.

@@

Another factor to consider is the design of achievement awards and points.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play  \cite{wbsnipes:Hamari2011Framework} .  How those achievements are awarded to players, and the concept of levels of achievements.  The article contains a warning that assigning achievements to tasks reduces intrinsic motivation because autonomy is reduced by the achievements.   In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.  When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.  Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{wbsnipes:Hamari2011Framework} making the game challenging for beginners and more experienced developers.  Designing the points and levels such that above average developers can complete all levels during the study period is ideal.

@@

\section{Related Work}
%update and expand

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influcence them to change their practices.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et. al. study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommender algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations. Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
